### 2009-06-01 paper.tex
Edits

Below we describe a method to eliminate null cycles from
the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
We then present multiple-sequence versions of the CYK, Inside and Outside algorithms.

As an aside: while some sort of null-cycle elimination is often required in order to deal with cyclic dependencies, there are several ways to accomplish this other than the algorithm presented here.
A simpler approach (that only works for the CYK algorithm) appears in the computational linguistics literature \cite{GecsegSteinby97}.
We have also developed a heuristic for CYK that simply ignores null cycles, as well as an iterative approximation that loops several times over cyclically-dependent cells of the DP matrix
until the estimate starts to converge. For conciseness, we have omitted descriptions of these methods, presenting only the exact elimination algorithm.

@@

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model construction.
While the singlet and branch transducers are relatively simple---the
singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
and 2 bifurcation states and the branch transducer, shown in
\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
states---the ensemble model of three extant sequences is very complex.
The naive exponential upper-bound gives a maximal state space of size
$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
uninformed search algorithm, we determined that there are 287
accessible states and 686 possible transitions between these states
(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
with the exponential calculation).  After performing the
transformations described in \secname{Exact elimination of null cycles in SCFGs} to
eliminate useless windback states, the ensemble model has a reduced
state space with 230 states, albeit at the cost of extra transitions,
bring the total to 1,789 transitions (here we are trading reduced
memory complexity, which is linear in the number of states, for
increased time complexity, which is linear in the number of
transitions).  Note that both before and after the reduction in
complexity, the total number of states and transitions are less than
the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
 Nonetheless, the extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.  Dataset S1 gives the state space of the composed
model constructed by the search algorithm and Dataset S2 the reduced
model after eliminating windback states; both are in Graphviz format for
visualization and show the state of the singlet transducer generating
ancestral sequence as well as the states of the branch transducers
generating observed sequences.

>>>>

Below we describe a method to eliminate null cycles from
the ensemble model by transforming any SCFG to an equivalent acyclic SCFG.
We then present multiple-sequence versions of the CYK, Inside and Outside algorithms.

While some sort of null-cycle elimination is often required in order
to deal with cyclic dependencies, there are several ways to accomplish
this other than the algorithm presented below.  A simpler approach
(that only works for the CYK algorithm) appears in the computational
linguistics literature \cite{GecsegSteinby97}.  We have also developed
a heuristic for CYK that simply ignores null cycles as well as an
iterative approximation that loops several times over
cyclically-dependent cells of the DP matrix until the estimate starts
to converge. For conciseness, we have omitted descriptions of these
methods, presenting only the exact elimination algorithm.

@@

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model construction.
While the singlet and branch transducers are relatively simple---the
singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
and 2 bifurcation states and the branch transducer, shown in
\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
states---the ensemble model of three extant sequences is very complex.
The naive exponential upper-bound gives a maximal state space of size
$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
uninformed search algorithm, we determined that there are 287
accessible states and 686 possible transitions between these states
(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
with the exponential calculation).  After performing the
transformations described in \secname{Exact elimination of null cycles in SCFGs} to
eliminate useless windback states, the ensemble model has a reduced
state space with 230 states, albeit at the cost of extra transitions,
bring the total to 1,789 transitions (here we are trading reduced
memory complexity, which is linear in the number of states, for
increased time complexity, which is linear in the number of
transitions).  Note that both before and after the reduction in
complexity, the total number of states and transitions are less than
the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
 Nonetheless, the extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.  Dataset S1 gives the state space of the ensemble
model constructed by the search algorithm and Dataset S2 the reduced
model after eliminating windback states; both are in Graphviz format for
visualization and show the state of the singlet transducer generating
ancestral sequence as well as the states of the branch transducers
generating observed sequences.
